Binning Strategies and Related Loss for Large Data

### 2015-02-18 Paper.Rnw
Standard binning reorder: define general case with bins specified by first center and binwidths, not data range special case

\subsection{Standard Rectangular Binning Algorithm}
We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner. Let $x_{(i)}$ correspond to the ordered values of $X$, where $i=1, ..., n$ and $n$ is the number of total observations. Then $x_{(1)}$ and $x_{(n)}$ then denote the observed minimum and the maximum of $X$, respectively.

We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$. Binning can be determined by the number of bins, $N_X$, which then dictates the bin width, $\omega_X$. For standard binning the relationship between these two parameters is  given as 
\[
\omega_X = (x_{(n)} - x_{(1)})/N_X
\]

Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_i \in \{ x_j^\ast \}, j \in \{1, ..., N_X\}$, i.e. the binning function   $b_X(.) : x_i \rightarrow x^\ast_i$ is defined as 
% \begin{center}
% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
% \vspace{.03}
% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
% \end{center}
\begin{eqnarray}\label{standbin}
 x^\ast_i = b_X(x_i) = \left\{\begin{array}{ll} 
    x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor & \quad \forall x_i < x_{(n)} \\
    x_{(n)} - \omega_x/2 & \quad \forall x_i = x_{(n)}
  \end{array}\right.
\end{eqnarray}

where $x_{(1)}^\ast = x_{(1)} + \omega_x/2$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ represents the number of bins centers that lie between $x_{(1)}$ and $x_i$.  Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between the bin centers for $x_{(1)}$ and $x_i$ to the known bin center for $x_{(1)}$.\\

Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width. A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins. This strategy would still fit into the framework of assigning values to the closest bin center.

@@

@@

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@
An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties. A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
  \end{array}\right.
\end{eqnarray}

>>>>

\subsection{Standard Rectangular Binning Algorithm}
We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner. We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.  To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$.  Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e. the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
% \begin{center}
% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
% \vspace{.03}
% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
% \end{center}
\begin{eqnarray}\label{standbin}
b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
\end{eqnarray}

where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.  Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.

A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.  To do so we use the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively. The first bin would then be centered at $x_{(1)} + \omega_X/2$.

%Note that the binning function above amounts to dividing up the range of data by a specified number of bins of equal width. A more flexible binning algorithm may be achieved if we allow the specification of the first bin center and bin width instead of the number of bins. This strategy would still fit into the framework of assigning values to the closest bin center.

\subsection{Random Binning Algorithm}

An alternative to the standard binning processes discussed above, is the random binning algorithm which results in statistically nice properties. A random binning algorithm utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability depending on the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } p = {(x^\ast_{j+1} - x_i)}/\omega_X \\
  x^\ast_{j+1} &\text{ with probability } 1-p = {(x_i - x^\ast_{j})}/\omega_X
  \end{array}\right.
\end{eqnarray}

\subsection{Extensions to Rectangular Binning}

@@

<<RandomBinDemo,echo=F,include=FALSE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

dnew$id <- interaction(dnew[,1:(nl)])
idx <- which(dnew$x%in%c(5, 15) & dnew$y==25)
d.rand$bin <- dnew$x%in%c(5, 15) & dnew$y==25
d.rand$a <- .33 + .66*d.rand$bin
d.rand$xbin <- NA
d.rand$ybin <- NA
d.rand$xbin[idx] <- dnew$x[idx]
d.rand$ybin[idx] <- 25
# res <- ddply(dnew, .(id), summarize, n=length(id), fsum=sum(Freq))
qplot(data=d.rand[,], x=x, y=y, alpha=a,  geom="point", xlim=c(-20, 40), ylim=c(0, 50)) +
  annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), fill="blue", alpha=.25) +
  annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), fill="green", alpha=.25) +
  scale_alpha(range=c(.5, 1), guide=FALSE) +
  scale_colour_identity(guide=FALSE) + 
  scale_shape(solid=T,guide=FALSE) +
  theme_bw() +
  annotate("point", x=rep(seq(-15, 35, 10),each=7), 
         y=rep(seq(-5, 55, 10), times=6), 
         shape='+', size=8, colour="gray90") + 
  geom_point(data=d.rand[which(!is.na(d.rand$xbin)),], 
             aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), size=I(3), aes.inherit=FALSE) + 
  scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
  scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
  annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + ylim(c(0, 50))
ggsave("./figure/RandomBinningDataPoints2.pdf",width=5, height=5)
@

@@
